---
title: "Antibiotics"
author: "(Your name here)"
date: 2020-
output:
  github_document
editor_options: 
  markdown: 
    wrap: 72
---

*Purpose*: Creating effective data visualizations is an *iterative*
process; very rarely will the first graph you make be the most
effective. The most effective thing you can do to be successful in this
iterative process is to *try multiple graphs* of the same data.

Furthermore, judging the effectiveness of a visual is completely
dependent on *the question you are trying to answer*. A visual that is
totally ineffective for one question may be perfect for answering a
different question.

In this challenge, you will practice *iterating* on data visualization,
and will anchor the *assessment* of your visuals using two different
questions.

*Note*: Please complete your initial visual design **alone**. Work on
both of your graphs alone, and save a version to your repo *before*
coming together with your team. This way you can all bring a diversity
of ideas to the table!

<!-- include-rubric -->

# Grading Rubric

<!-- -------------------------------------------------- -->

Unlike exercises, **challenges will be graded**. The following rubrics
define how you will be graded, both on an individual and team basis.

## Individual

<!-- ------------------------- -->

| Category    | Needs Improvement                                                                                                | Satisfactory                                                                                                               |
|------------------|-----------------------------|-------------------------|
| Effort      | Some task **q**'s left unattempted                                                                               | All task **q**'s attempted                                                                                                 |
| Observed    | Did not document observations, or observations incorrect                                                         | Documented correct observations based on analysis                                                                          |
| Supported   | Some observations not clearly supported by analysis                                                              | All observations clearly supported by analysis (table, graph, etc.)                                                        |
| Assessed    | Observations include claims not supported by the data, or reflect a level of certainty not warranted by the data | Observations are appropriately qualified by the quality & relevance of the data and (in)conclusiveness of the support      |
| Specified   | Uses the phrase "more data are necessary" without clarification                                                  | Any statement that "more data are necessary" specifies which *specific* data are needed to answer what *specific* question |
| Code Styled | Violations of the [style guide](https://style.tidyverse.org/) hinder readability                                 | Code sufficiently close to the [style guide](https://style.tidyverse.org/)                                                 |

## Due Date

<!-- ------------------------- -->

All the deliverables stated in the rubrics above are due **at midnight**
before the day of the class discussion of the challenge. See the
[Syllabus](https://docs.google.com/document/d/1qeP6DUS8Djq_A0HMllMqsSqX3a9dbcx1/edit?usp=sharing&ouid=110386251748498665069&rtpof=true&sd=true)
for more information.

```{r setup}
library(tidyverse)
library(ggrepel)
```

*Background*: The data[1] we study in this challenge report the
[*minimum inhibitory
concentration*](https://en.wikipedia.org/wiki/Minimum_inhibitory_concentration)
(MIC) of three drugs for different bacteria. The smaller the MIC for a
given drug and bacteria pair, the more practical the drug is for
treating that particular bacteria. An MIC value of *at most* 0.1 is
considered necessary for treating human patients.

These data report MIC values for three antibiotics---penicillin,
streptomycin, and neomycin---on 16 bacteria. Bacteria are categorized
into a genus based on a number of features, including their resistance
to antibiotics.

```{r load}
## NOTE: If you extracted all challenges to the same location,
## you shouldn't have to change this filename
filename <- "./data/antibiotics.csv"

## Load the data
df_antibiotics <- read_csv(filename)
df_antibiotics %>% knitr::kable()
```

# Visualization

<!-- -------------------------------------------------- -->

### **q1** Prototype 5 visuals

To start, construct **5 qualitatively different visualizations of the
data** `df_antibiotics`. These **cannot** be simple variations on the
same graph; for instance, if two of your visuals could be made identical
by calling `coord_flip()`, then these are *not* qualitatively different.

For all five of the visuals, you must show information on *all 16
bacteria*. For the first two visuals, you must *show all variables*.

*Hint 1*: Try working quickly on this part; come up with a bunch of
ideas, and don't fixate on any one idea for too long. You will have a
chance to refine later in this challenge.

*Hint 2*: The data `df_antibiotics` are in a *wide* format; it may be
helpful to `pivot_longer()` the data to make certain visuals easier to
construct.

#### Visual 1 (All variables)

In this visual you must show *all three* effectiveness values for *all
16 bacteria*. You must also show whether or not each bacterium is Gram
positive or negative.

```{r q1.1}
# WRITE YOUR CODE HERE
df_longer <- df_antibiotics %>%
  pivot_longer(
    names_to = "antibiotic",
    values_to = "MIC",
    cols = c(penicillin, streptomycin, neomycin)
  ) 
df_longer %>%
    ggplot() +
    geom_dotplot(mapping = aes(x = gram, y = MIC, fill = bacteria), binaxis = "y", 
               dotsize = 0.2, binwidth = 1) +
    scale_y_log10() +
    facet_wrap(~ antibiotic)
```

#### Visual 2 (All variables)

In this visual you must show *all three* effectiveness values for *all
16 bacteria*. You must also show whether or not each bacterium is Gram
positive or negative.

Note that your visual must be *qualitatively different* from *all* of
your other visuals.

```{r q1.2}
# WRITE YOUR CODE HERE
df_longer %>%
  ggplot(aes(shape = gram)) +
  geom_point(mapping = aes(x = bacteria, y = MIC, color = antibiotic)) +
  geom_point(mapping = aes(x = bacteria, y = 0.1), color = "black") +
  scale_y_log10() +
  labs(y="MIC") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
  
```

#### Visual 3 (Some variables)

In this visual you may show a *subset* of the variables (`penicillin`,
`streptomycin`, `neomycin`, `gram`), but you must still show *all 16
bacteria*.

Note that your visual mustsf be *qualitatively different* from *all* of
your other visuals.

```{r q1.3}
# WRITE YOUR CODE HERE

df_antibiotics %>%
  mutate(treatment = ifelse((penicillin <= 0.1) | (streptomycin <= 0.1) | (neomycin <= 0.1), "yes", "no")) %>%
  mutate(count = (penicillin <= 0.1) + (streptomycin <= 0.1) + (neomycin <= 0.1)) %>%
  pivot_longer(
    names_to = "antibiotic",
    values_to = "MIC",
    cols = c(penicillin, streptomycin, neomycin)
  ) %>%
  ggplot() +
  geom_col(mapping = aes(x = treatment, y = count)) +
  facet_wrap(~ bacteria)
```

#### Visual 4 (Some variables)

In this visual you may show a *subset* of the variables (`penicillin`,
`streptomycin`, `neomycin`, `gram`), but you must still show *all 16
bacteria*.

Note that your visual must be *qualitatively different* from *all* of
your other visuals.

```{r q1.4}
# WRITE YOUR CODE HERE
df_longer %>%
  ggplot() +
  geom_tile(mapping = aes(x = antibiotic, y = bacteria, fill = log10(MIC)))
  
df_longer
```

#### Visual 5 (Some variables)

In this visual you may show a *subset* of the variables (`penicillin`,
`streptomycin`, `neomycin`, `gram`), but you must still show *all 16
bacteria*.

Note that your visual must be *qualitatively different* from *all* of
your other visuals.

```{r q1.5}
# WRITE YOUR CODE HERE
df_longer %>%
    filter(MIC <= 0.1) %>%
      ggplot() +
      geom_point(mapping = aes(x = bacteria, y = MIC, color = antibiotic)) +
      theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))

```

### **q2** Assess your visuals

There are **two questions** below; use your five visuals to help answer
both Guiding Questions. Note that you must also identify which of your
five visuals were most helpful in answering the questions.

*Hint 1*: It's possible that *none* of your visuals is effective in
answering the questions below. You may need to revise one or more of
your visuals to answer the questions below!

*Hint 2*: It's **highly unlikely** that the same visual is the most
effective at helping answer both guiding questions. **Use this as an
opportunity to think about why this is.**

#### Guiding Question 1

> How do the three antibiotics vary in their effectiveness against
> bacteria of different genera and Gram stain?

*Observations* -

What is your response to the question above?

-   Penicillin seems to be very effective in treating bacteria in the
    genus of staphylococcus or streptococcus (all gram positive) as the
    MIC of penicillin is sufficiently low to be effective in human
    treatment. It's rather impractical for other strains, though it does
    inhibit their growth eventually.

-   Neomycin is effective against strains of salmonella as well as the
    brucella strain. These are gram negative bacteria. It also does
    reasonably well against three gram positive bacteria: one bacillus,
    one streptococcus, and one staphylococcus. In general, the MIC tends
    to be smaller and seems to have less spread for gram negative
    bacteria.

-   Streptomycin seems to be effective and practical for only four
    genera, proteus (gram negative), bacillus (gram positive), and the
    two staphylococci in the dataset (both gram positive). This seems to
    indicate that practicality favors streptomycin use against gram
    positive bacteria, although the eye test indicates that on average
    the gram negative bacteria might have lower MIC.

Which of your visuals above (1 through 5) is **most effective** at
helping to answer this question?

-   The dot plot (1) was easily most effective at answering this
    question.

Why?

-   It was easy to have a visualization where data is separated between
    gram positive and negative, whereas it was only indicated by shape
    with others, which would've been harder to read.

-   Having all bacteria colored was nice, especially given that there
    are few enough bacteria in the dataset where colors are
    distinguishable. It was also fortunate that their order is one such
    that similar colors are from similar strains (e.g. staphylococci).

#### Guiding Question 2

In 1974 *Diplococcus pneumoniae* was renamed *Streptococcus pneumoniae*,
and in 1984 *Streptococcus fecalis* was renamed *Enterococcus fecalis*
[2].

> Why was *Diplococcus pneumoniae* was renamed *Streptococcus
> pneumoniae*?

*Observations* - What is your response to the question above?

-   Against various antibiotics, they receive very similar MIC ratings.

-   They are all gram positive.

Which of your visuals above (1 through 5) is **most effective** at
helping to answer this question?

-   My scatterplot (2) was most effective at answering this question.

Why?

-   It contains information about both their MIC and gram grades.
    Visualizations 3 and 5 target the number of/kind of antibiotics that
    are worthy for human treatment, and were designed to answer
    questions more in the line of "how practical is "x" antibiotic
    against "y" bacteria. Visualization 1 makes it difficult to see
    diplococcus pneumoniae since it is so close to streptococci, but it
    it were distinguishable then perhaps visualization 1 would be in
    contention. Visualization 4 is certainly a winner, because the
    heatmap's pattern for diplococcus pneumoniae echoes what can be seen
    for two of the three streptococci very similarly. On the other hand,
    it didn't answer the question of whether they were gram positive or
    gram negative, so I decided that visualization 2 helped me answer
    the question best.

# References

<!-- -------------------------------------------------- -->

[1] Neomycin in skin infections: A new topical antibiotic with wide
antibacterial range and rarely sensitizing. Scope. 1951;3(5):4-7.

[2] Wainer and Lysen, "That's Funny..." *American Scientist* (2009)
[link](https://www.americanscientist.org/article/thats-funny)
